The investment strategy based on behavior of artificial earthworm for use in algotrading

### 2014-05-11 Article/Article20140406Granada/LatexArt/article_main.tex
worm->earthworm

Strategy presented in this paper is an example of a solution that can be used fully automatically or with assistance of investor who makes the decision manually. Using such softwaremakes sense only if there are reasonable premises of achieving the success. These conditions can be achieved through inductive thinking \cite{provost,KleskWilinski}, involving the assumption that multiple validation of decision on data from the past substantiates the correctness of the decision in the future. If a certain set of rules in the strategy used to seek the pattern, leads to the success then these rules are a pattern that is encouraged for repeated use. If repeated use of mentioned pattern results in more successes than failures, then the basic premise to consider the use of the pattern is rational \cite{bishop2006pattern,KleskWilinski}. Unfortunately, changing market is causing the pattern to become outdated therefore it is necessary to constantly update it. Finding historically effective patterns that could be also effective in a data space not used for their determination (to identify these patterns) is a major challenge for machine learning which is applied in the paper.
The article is organized as follows: at first authors describe an authored strategy based on an entirely innovative concept of an artificial worm. Subsequently extensive studies of effectiveness of presented strategy in terms of prediction and the aspect of investment are carried out. At the end the conclusions resulting from the study are presented and also cited bibliographic sources are included.

\section{Strategy characteristics}
In presented strategy, a metaphorical procedure is often used to explain heuristic algorithms of artificial intelligence. Said procedure is a reference to an inspiration derived from the nature.\\

Let's consider an artificial earthworm, which behaves like its natural inspirer. Let this artificial worm have a length of $p1$ candles (vectors representing the knowledge about events occurring within the contractual part of the time series). Let her movement come down to the successive ,,eating'' one candle and simultaneous ,,discarding'' of one candle (the last one) --- always having $p1$ candles. Thus the earthworm moves incrementally at length of time called the period of the candle. $i-th$ candle is  defined as a row of data of values  $O_i$, $H_i$, $L_i$, $C_i$ (Open, High, Low and Close). These values represent (in order): the value at the beginning of this period, the maximum value within the period, the minimum and the closing values - at the end of the candle.\\
Let's assume that candle can have color (often in a number of brokerage platforms, this approach is used). Therefore candles may be red (when the opening of the candle is higher than its closing $O_i>C_i$) and green (when the situation is opposite $O_i<C_i$).\\
Let's also assume --- it is a primary strategy rule - that when an artificial earthworm has a significant dominance of red candles, a short position is opened or continued. When earthworm has the significant dominance of green candles, a long position is opened or continued. Imprecise term ,,significant'' is defined by the use of the second parameter $p2$ according to the rules:
\begin{equation}
\text{If }  Lr > p1/2 -p2 \text{ then } short;
\end{equation}
opening of short position will be suggested;
\begin{equation}
\text{If }  Lg > p1/2 -p2 \text{ then } long;
\end{equation}
opening of long position will be suggested;
\begin{equation}
\text{If }  Lg < p1/2 -p2 \text{ and } Lr< p1/2-p2 \text{ then } none;
\label{eq:eq3}
\end{equation}
the continuation of state will be suggested (position or no opened position).\\

In the rules above $Lr$ and $Lg$ are the number of candles, respectively ,,red'' and ,,green'' in the worm of $p1$ length, wherein $p1 = Lr + Lg$.\\
When dominance determined by another parameter $p2$ is undecided, as in (\ref{eq:eq3}), then worm takes the color white and the position will not be opened. When the worm continues to open position, investor does not incur costs. On many platforms, also in literature, symbolic marking of growing candles in white and decreasing in black is used. In this article, due to the occurrence of three values, green-red-white colors are used.\\

Earthworm is therefore supplied with artificial intellect, which allows the accumulation of knowledge and decision-making:
\begin{itemize}
\item right after closing of the current candle, worm ,,swallows'' the candle and ,,knows'' what color is it;
\item earthworm knows how many candles it is composed with;
\item after absorbing and discarding one candle, worm knows how many candles inside it are red or green and how to determine its status --- red, green or white;
\item knowing its status, earthworm knows how to make a decision about opening or continuing position; the opening of a position (long or short) means the simultaneous closing of the current position (opposite) if such position is opened;
\item earthworm observes cumulative profit curve and if it  notices unacceptable drawdown, then opened position is changed to the opposite;
\item unacceptable drawdown is defined by two other parameters $p3$ and $p4$. $p3$ is the number of previous candles, which sets a reference point in time series, in which the value of the accumulated profit for $i-p3$ is checked for comparison with the current value of the accumulated profit. $p4$ is permissible decrease of cumulative profit. If it is exceeded, the $SL$ (Stop Loss --- closes the currently open position) mechanism is activated;
\item worm also observes current course changes during  each next candle and reacts to unprofitable changes. If the current exchange rate causes a loss on the opened positions greater than $p5$ then the $SL$ mechanism is triggered --- the position is closed and till the end of the candle worm remains in the state without an open position. $p5$ is therefore the fifth strategy parameter. 
\end{itemize}

These features of the worm's intelligence may be extended, however, for this work they were limited to the foregoing. \\

@@

Artificial earthworm executes the following algorithm.\\
The strategy is built on the principle of classical machine learning. Before starting to run the algorithm, on data located outside the environment that investor knows (like on the current market segments --- out-of-sample), it is necessary to ,,teach''  the values of parameters. This principle is presented in Figure \ref{fig:fig2}, which shows examples of learning periods and following them testing periods. Data from testing period was not used during learning.\\
\begin{figure}[ht]
\centering
\includegraphics[width = 0.6\textwidth]{figures/rys2.png}
\caption{An example of the arrangement of learning and testing periods of strategy parameters in time series.}
\label{fig:fig2}
\end{figure}
\FloatBarrier
\vspace{-1em}
This repeatable algorithm, located in both periods - learning and testing, has the following structure. In learning period the optimal value for five parameters $p1$, $p2$, ..., $p5$ was determined. In test period these designated values of parameters were used to perform the algorithm in the new environment. In each $i-th$ step (candle) of time series following activities were performed:
\begin{itemize}
\item For each $i = p1, p1 +1, p1 +2, ..., I$ and for $j = 1, ..., p1$ check whether the candle located at a distance $i-j$ backwards from the current candle is:\\
green when $O_{i-j} <C_{i-j}$, \\
red when $O_{i-j}> C_{i-j}$, \\
where $p1$ --- is the first parameter which is the length of the artificial worm --- the number of candles of different colors that make up the ,,body'' of the worm; \\
$O_{i-j}$ --- is the opening value of (i-j)-th candle;\\ 
$C_{i-j}$ --- is the closing value of (i-j)-th candle; \\
$I$ --- is the length of the test window, after which it is necessary to change the parameters values to the new, learned in the machine learning phase;\\
\item In each $i-th$ step, the loop $j = 1, ..., p1$ is performed in which a number of green $Lg$, red $Lr$ candles is calculated so the sum of $Lg$ and $Lr$ is $p1$ (to avoid the situation when $O_i = C_i$, it is assigned arbitrarily to one of the candles sets, for example green);
\item A condition for the existence of a $Rg$ recommendation to open a long position (or to continue it if it is open) is checked. This condition is arbitrarily adopted (one can of course consider different one): 
\begin{equation}
\text{If } Lg> p1/2-p2 \text{ then } Rg=1
\end{equation}

@@

In formulas above there is a parameter $p2$. It is the second parameter of the strategy which is an arbitrarily separated part of the worm, which depending on size causes the direction of investment not to change. Large $p2$ causes long positions to change into short positions (or vice-versa) rarely. Small value of $p2$ leads to a higher frequency of these changes. The optimum value of $p2$ is chosen similarly like $p1$.
\begin{itemize}
\item 	When opening long position, profit after one candle (this does not mean that position was closed after a single candle) can be calculated as:
\begin{equation}
Z_i = C_{i+1} - C_i - spread;
\end{equation}
where the $spread = 0$, when opened long position is continued; \\
and $spread = cost$, when long position was opened after closing or opening a short position in case when there was no open position;\\
$cost$ is a variable that depends on the particular brokerage platform that was used for testing the strategy; in this case, for the currency pair EURUSD $cost = 0.00016$ was assumed and for the currency pair GBPUSD $cost = 0.00028$.
\item When short position is opened, profit after one candle (it does not mean that position is closed) can be calculated as: 
\begin{equation}
Z_i = C_i-C_{i+1}- spread;
\end{equation}
where the $spread = 0$, when opened short position is continued;\\ 
and $spread = cost$, when short position was opened after closing or opening a long position in case when there was no open position.
\end{itemize}

\begin{itemize}
\item If as the result of the test the number of red and green candles in the body of the worm had not obtained any dominance from the color that would recommend to open defined position (worm was white), it is assumed to continue the position, which is open, without any cost of the transaction (with the value of $spread = 0$). If at the beginning of the candle, at the time of checking the recommendation, there was no open position then it can be assumed to continue without any position.
\item Each of the profits calculated above is verified in terms of exceeding the allowable local position drawdown (within one candle). This additional risk management is widely used in automated trading and is called limiting losses --- Stop Loss ($SL$). In this algorithm, $SL$ was introduced as the fifth parameter of the strategy --- $p5$.
\end{itemize}

\noindent Presented strategy was tested if:
\begin{itemize}
\item in the case of opening at the beginning of the $i-th$ candle or continuation of a long position presented condition did not occur:\\
\begin{equation}
\text{If } L_i-O_i<SL \text{ then } Z_i=-SL
\end{equation}
where $L_i$ --– the lowest rate value in the $i-th$ candle; $SL$ --– absolute value Stop Loss appearing as $p5$ parameter;
\item in the case of opening at the beginning of the $i-th$ candle or continuation of a short position presented condition did not occur:
\begin{equation}
\text{If } O_i-H_i<SL \text{ then } Z_i=-SL
\end{equation}
where $H_i$ --– the highest rate value in the $i-th$ candle.
\end{itemize}
This way $p5$ parameter was used for securing in strategy the strong local changes in exchange rate within a single candle.\\
The meaning of $p5$ parameter to the behavior of the worm and the occurrence of the closing position is explained in Figure \ref{fig:fig5}.
\begin{figure}[h!]
\centering
\includegraphics[width = 0.5\textwidth]{figures/rys5.png}
\caption{Method of calculating the profit (here: loss) in the case of necessity to use $p5$ parameter actuating mechanism $SL$ for long positions.}
\label{fig:fig5}
\end{figure}
\FloatBarrier

@@

In the article, implementation for MetaTrader4 has covered only the testing part of the algorithm, using fixed set of parameters that have been previously computed in MATLAB in the learning part of execution for every test period. It is important to note that implementing full algorithm, along with the learning part in MQL4, is obviously possible, however in the article the main objective was to confirm that the artificial worm can run on a real market with results similar to those achieved in MATLAB.\\

@@

Results of the algorithm execution in both environments for both currency pairs are presented in Table \ref{tab:tab5}.
\begin{table}[h!]
\centering
\caption{Results of the algorithm execution in both environments for both currency pairs and various parameters values.}
\label{tab:tab5}
\begin{tabular}{|p{0.19\textwidth}|p{0.19\textwidth}|p{0.19\textwidth}|p{0.19\textwidth}|p{0.19\textwidth}|} \hline
 &  Total net profit in MetaTrader [EUR]	 &  Total net profit in MATLAB 
[EUR]	 &  Difference [EUR]	 &  Difference [\%]\\ \hline
\multicolumn{5}{|c|}{GBPUSD in time period 2013.01.09 - 2013.12.30}\\ \hline
Fixed $p1$-$p5$	 & -8 823.76 & 	-7 987.00 & 	836.76 & 	10.48\\
Dynamic $p1$-$p5$	 & 6 057.82	 & 5 555.00	 & 502.82 & 	9.05\\ \hline
\multicolumn{5}{|c|}{EURUSD in time period 2012.01.20 - 2013.01.01}\\ \hline
Fixed $p1$-$p5$	 & 2 305,18 & 	2 695,00	 & 389,82	 & 14,46\\
Dynamic $p1$-$p5$ & 	13 669,40 & 	17 260,00 & 	3590,60	 & 20,80\\ \hline
\end{tabular}
\end{table}
\FloatBarrier
\begin{figure}[h!]
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width = 0.9\textwidth]{figures/rys12.png}
\subcaption{Balance (relative value) in MATLAB.}
\label{fig:fig12}
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width = \textwidth]{figures/rys13.png}
\subcaption{Balance (absolute value) in MetaTrader.}
\label{fig:fig13}
\end{minipage}
\caption{Balance for GBPUSD [1h] in period 2013.01.09 - 2013.12.30 with fixed parameters $p1$-$p5$.}
\end{figure}
\FloatBarrier
\vspace{-1em}
It is important to note that every first chart from each pair above was done in MATLAB and every second chart was done in MetaTrader. In MATLAB chart context, every point of abscissa axis represents testing period, while on ordinate axis there is a result of testing period (relative value), so that it can be assumed that chart was done in time domain. On the other hand, every abscissa on MetaTrader chart represents each opening of position, which means that chart is not linear related with time (however in cases presented in the article it is really similar to time domain), so that chart begins with first position opening and it is denoised in comparison to MATLAB chart as every line are determined by opening and closing position, no matter how long this position has lasted. Despite those differences, correlation between MATLAB and MetaTrader charts for the artificial worm algorithm executions is indisputable.
\begin{figure}[h!]
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width = 0.9\textwidth]{figures/rys14.png}
\subcaption{Balance (relative value) in MATLAB.}
\label{fig:fig14}
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width = \textwidth]{figures/rys15.png}
\subcaption{Balance (absolute value) in MetaTrader.}
\label{fig:fig15}
\end{minipage}
\caption{Balance for GBPUSD [1h] in period 2013.01.09 - 2013.12.30 with dynamic parameters $p1$-$p5$.}
\end{figure}
\FloatBarrier
\vspace{-1em}

>>>>

Strategy presented in this paper is an example of a solution that can be used fully automatically or with assistance of investor who makes the decision manually. Using such softwaremakes sense only if there are reasonable premises of achieving the success. These conditions can be achieved through inductive thinking \cite{provost,KleskWilinski}, involving the assumption that multiple validation of decision on data from the past substantiates the correctness of the decision in the future. If a certain set of rules in the strategy used to seek the pattern, leads to the success then these rules are a pattern that is encouraged for repeated use. If repeated use of mentioned pattern results in more successes than failures, then the basic premise to consider the use of the pattern is rational \cite{bishop2006pattern,KleskWilinski}. Unfortunately, changing market is causing the pattern to become outdated therefore it is necessary to constantly update it. Finding historically effective patterns that could be also effective in a data space not used for their determination (to identify these patterns) is a major challenge for machine learning which is applied in the paper.
The article is organized as follows: at first authors describe an authored strategy based on an entirely innovative concept of an artificial earthworm. Subsequently extensive studies of effectiveness of presented strategy in terms of prediction and the aspect of investment are carried out. At the end the conclusions resulting from the study are presented and also cited bibliographic sources are included.

\section{Strategy characteristics}
In presented strategy, a metaphorical procedure is often used to explain heuristic algorithms of artificial intelligence. The procedure is a reference to an inspiration derived from the nature.\\

Let's consider an artificial earthworm, which behaves like its natural inspirer. Let this artificial earthworm have a length of $p1$ candles (vectors representing the knowledge about events occurring within the contractual part of the time series). Let its movement come down to the successive ,,eating'' one candle and simultaneous ,,discarding'' of one candle (the last one) --- always having $p1$ candles. Thus the earthworm moves incrementally at length of time called the period of the candle. $i-th$ candle is  defined as a row of data of values  $O_i$, $H_i$, $L_i$, $C_i$ (Open, High, Low and Close). These values represent (in order): the value at the beginning of this period, the maximum value within the period, the minimum and the closing values - at the end of the candle.\\
Let's assume that candle have a color (often in a number of brokerage platforms such approach is used). Therefore candles may be red (when the opening of the candle is higher than its closing $O_i>C_i$) and green (when the situation is opposite $O_i<C_i$).\\
Let's also assume --- it is a primary strategy rule - that when the artificial earthworm has a significant dominance of red candles, a short position is opened or continued. When earthworm has the significant dominance of green candles, a long position is opened or continued. Imprecise term ,,significant'' is defined by the use of the second parameter $p2$ according to the rules:
\begin{equation}
\text{If }  Lr > p1/2 -p2 \text{ then } short;
\end{equation}
opening of short position will be suggested;
\begin{equation}
\text{If }  Lg > p1/2 -p2 \text{ then } long;
\end{equation}
opening of long position will be suggested;
\begin{equation}
\text{If }  Lg < p1/2 -p2 \text{ and } Lr< p1/2-p2 \text{ then } none;
\label{eq:eq3}
\end{equation}
the continuation of state will be suggested (position or no opened position).\\

In the rules above $Lr$ and $Lg$ are the number of candles, respectively ,,red'' and ,,green'' in the earthworm of $p1$ length, wherein $p1 = Lr + Lg$.\\
When dominance determined by another parameter $p2$ is undecided, as in (\ref{eq:eq3}), then earthworm takes the color white and the position will not be opened. When the earthworm continues to open position, investor does not incur costs. On many platforms, also in literature, symbolic marking of growing candles in white and decreasing in black is used. In this article, due to the occurrence of three values, green-red-white colors are used.\\

Earthworm is therefore supplied with artificial intellect, which allows the accumulation of knowledge and decision-making:
\begin{itemize}
\item right after closing of the current candle, earthworm ,,swallows'' the candle and ,,knows'' what color is it;
\item earthworm knows how many candles it is composed with;
\item after absorbing and discarding one candle, earthworm knows how many candles inside it are red or green and how to determine its status --- red, green or white;
\item knowing its status, earthworm knows how to make a decision about opening or continuing position; the opening of a position (long or short) means the simultaneous closing of the current position (opposite) if such position is opened;
\item earthworm observes cumulative profit curve and if it  notices unacceptable drawdown, then opened position is changed to the opposite;
\item unacceptable drawdown is defined by two other parameters $p3$ and $p4$. $p3$ is the number of previous candles, which sets a reference point in time series, in which the value of the accumulated profit for $i-p3$ is checked for comparison with the current value of the accumulated profit. $p4$ is permissible decrease of cumulative profit. If it is exceeded, the $SL$ (Stop Loss --- closes the currently open position) mechanism is activated;
\item earthworm also observes current course changes during  each next candle and reacts to unprofitable changes. If the current exchange rate causes a loss on the opened positions greater than $p5$ then the $SL$ mechanism is triggered --- the position is closed and till the end of the candle earthworm remains in the state without an open position. $p5$ is therefore the fifth strategy parameter. 
\end{itemize}

These features of the earthworm's intelligence may be extended, however, for this work they were limited to the foregoing. \\

@@

Artificial earthworm executes the following algorithm.\\
The strategy is built on the principle of classical machine learning. Before starting to run the algorithm, on data located outside the environment that investor knows (like on the current market segments --- out-of-sample), it is necessary to ,,teach''  the values of parameters. This principle is presented in Figure \ref{fig:fig2}, which shows examples of learning periods and following them testing periods. Data from testing period was not used during learning.\\
\begin{figure}[ht]
\centering
\includegraphics[width = 0.6\textwidth]{figures/rys2.png}
\caption{An example of the arrangement of learning and testing periods of strategy parameters in time series.}
\label{fig:fig2}
\end{figure}
\FloatBarrier
\vspace{-1em}
This repeatable algorithm, located in both periods - learning and testing, has the following structure. In learning period the optimal value for five parameters $p1$, $p2$, ..., $p5$ was determined. In test period these designated values of parameters were used to perform the algorithm in the new environment. In each $i-th$ step (candle) of time series following activities were performed:
\begin{itemize}
\item For each $i = p1, p1 +1, p1 +2, ..., I$ and for $j = 1, ..., p1$ check whether the candle located at a distance $i-j$ backwards from the current candle is:\\
green when $O_{i-j} <C_{i-j}$, \\
red when $O_{i-j}> C_{i-j}$, \\
where $p1$ --- is the first parameter which is the length of the artificial earthworm --- the number of candles of different colors that make up the ,,body'' of the earthworm; \\
$O_{i-j}$ --- is the opening value of (i-j)-th candle;\\ 
$C_{i-j}$ --- is the closing value of (i-j)-th candle; \\
$I$ --- is the length of the test window, after which it is necessary to change the parameters values to the new, learned in the machine learning phase;\\
\item In each $i-th$ step, the loop $j = 1, ..., p1$ is performed in which a number of green $Lg$, red $Lr$ candles is calculated so the sum of $Lg$ and $Lr$ is $p1$ (to avoid the situation when $O_i = C_i$, it is assigned arbitrarily to one of the candles sets, for example green);
\item A condition for the existence of a $Rg$ recommendation to open a long position (or to continue it if it is open) is checked. This condition is arbitrarily adopted (one can of course consider different one): 
\begin{equation}
\text{If } Lg> p1/2-p2 \text{ then } Rg=1
\end{equation}

@@

In formulas above there is a parameter $p2$. It is the second parameter of the strategy which is an arbitrarily separated part of the earthworm, which depending on size causes the direction of investment not to change. Large $p2$ causes long positions to change into short positions (or vice-versa) rarely. Small value of $p2$ leads to a higher frequency of these changes. The optimum value of $p2$ is chosen similarly like $p1$.
\begin{itemize}
\item 	When opening long position, profit after one candle (this does not mean that position was closed after a single candle) can be calculated as:
\begin{equation}
Z_i = C_{i+1} - C_i - spread;
\end{equation}
where the $spread = 0$, when opened long position is continued; \\
and $spread = cost$, when long position was opened after closing or opening a short position in case when there was no open position;\\
$cost$ is a variable that depends on the particular brokerage platform that was used for testing the strategy; in this case, for the currency pair EURUSD $cost = 0.00016$ was assumed and for the currency pair GBPUSD $cost = 0.00028$.
\item When short position is opened, profit after one candle (it does not mean that position is closed) can be calculated as: 
\begin{equation}
Z_i = C_i-C_{i+1}- spread;
\end{equation}
where the $spread = 0$, when opened short position is continued;\\ 
and $spread = cost$, when short position was opened after closing or opening a long position in case when there was no open position.
\end{itemize}

\begin{itemize}
\item If as the result of the test the number of red and green candles in the body of the earthworm had not obtained any dominance from the color that would recommend to open defined position (earthworm was white), it is assumed to continue the position, which is open, without any cost of the transaction (with the value of $spread = 0$). If at the beginning of the candle, at the time of checking the recommendation, there was no open position then it can be assumed to continue without any position.
\item Each of the profits calculated above is verified in terms of exceeding the allowable local position drawdown (within one candle). This additional risk management is widely used in automated trading and is called limiting losses --- Stop Loss ($SL$). In this algorithm, $SL$ was introduced as the fifth parameter of the strategy --- $p5$.
\end{itemize}

\noindent Presented strategy was tested if:
\begin{itemize}
\item in the case of opening at the beginning of the $i-th$ candle or continuation of a long position presented condition did not occur:\\
\begin{equation}
\text{If } L_i-O_i<SL \text{ then } Z_i=-SL
\end{equation}
where $L_i$ --– the lowest rate value in the $i-th$ candle; $SL$ --– absolute value Stop Loss appearing as $p5$ parameter;
\item in the case of opening at the beginning of the $i-th$ candle or continuation of a short position presented condition did not occur:
\begin{equation}
\text{If } O_i-H_i<SL \text{ then } Z_i=-SL
\end{equation}
where $H_i$ --– the highest rate value in the $i-th$ candle.
\end{itemize}
This way $p5$ parameter was used for securing in strategy the strong local changes in exchange rate within a single candle.\\
The meaning of $p5$ parameter to the behavior of the earthworm and the occurrence of the closing position is explained in Figure \ref{fig:fig5}.
\begin{figure}[h!]
\centering
\includegraphics[width = 0.5\textwidth]{figures/rys5.png}
\caption{Method of calculating the profit (here: loss) in the case of necessity to use $p5$ parameter actuating mechanism $SL$ for long positions.}
\label{fig:fig5}
\end{figure}
\FloatBarrier

@@

In the article, implementation for MetaTrader4 has covered only the testing part of the algorithm, using fixed set of parameters that have been previously computed in MATLAB in the learning part of execution for every test period. It is important to note that implementing full algorithm, along with the learning part in MQL4, is obviously possible, however in the article the main objective was to confirm that the artificial earthworm can run on a real market with results similar to those achieved in MATLAB.\\

@@

Results of the algorithm execution in both environments for both currency pairs are presented in Table \ref{tab:tab5}.
\begin{table}[h!]
\centering
\caption{Results of the algorithm execution in both environments for both currency pairs and various parameters values.}
\label{tab:tab5}
\begin{tabular}{|p{0.19\textwidth}|p{0.19\textwidth}|p{0.19\textwidth}|p{0.19\textwidth}|p{0.19\textwidth}|} \hline
 &  Total net profit in MetaTrader [EUR]	 &  Total net profit in MATLAB 
[EUR]	 &  Difference [EUR]	 &  Difference [\%]\\ \hline
\multicolumn{5}{|c|}{GBPUSD in time period 2013.01.09 - 2013.12.30}\\ \hline
Fixed $p1$-$p5$	 & -8 823.76 & 	-7 987.00 & 	836.76 & 	10.48\\
Dynamic $p1$-$p5$	 & 6 057.82	 & 5 555.00	 & 502.82 & 	9.05\\ \hline
\multicolumn{5}{|c|}{EURUSD in time period 2012.01.20 - 2013.01.01}\\ \hline
Fixed $p1$-$p5$	 & 2 305,18 & 	2 695,00	 & 389,82	 & 14,46\\
Dynamic $p1$-$p5$ & 	13 669,40 & 	17 260,00 & 	3590,60	 & 20,80\\ \hline
\end{tabular}
\end{table}
\FloatBarrier
\begin{figure}[h!]
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width = 0.9\textwidth]{figures/rys12.png}
\subcaption{Balance (relative value) in MATLAB.}
\label{fig:fig12}
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width = \textwidth]{figures/rys13.png}
\subcaption{Balance (absolute value) in MetaTrader.}
\label{fig:fig13}
\end{minipage}
\caption{Balance for GBPUSD [1h] in period 2013.01.09 - 2013.12.30 with fixed parameters $p1$-$p5$.}
\end{figure}
\FloatBarrier
\vspace{-1em}
It is important to note that every first chart from each pair above was done in MATLAB and every second chart was done in MetaTrader. In MATLAB chart context, every point of abscissa axis represents testing period, while on ordinate axis there is a result of testing period (relative value), so that it can be assumed that chart was done in time domain. On the other hand, every abscissa on MetaTrader chart represents each opening of position, which means that chart is not linear related with time (however in cases presented in the article it is really similar to time domain), so that chart begins with first position opening and it is denoised in comparison to MATLAB chart as every line are determined by opening and closing position, no matter how long this position has lasted. Despite those differences, correlation between MATLAB and MetaTrader charts for the artificial earthworm algorithm executions is indisputable.
\begin{figure}[h!]
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width = 0.9\textwidth]{figures/rys14.png}
\subcaption{Balance (relative value) in MATLAB.}
\label{fig:fig14}
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width = \textwidth]{figures/rys15.png}
\subcaption{Balance (absolute value) in MetaTrader.}
\label{fig:fig15}
\end{minipage}
\caption{Balance for GBPUSD [1h] in period 2013.01.09 - 2013.12.30 with dynamic parameters $p1$-$p5$.}
\end{figure}
\FloatBarrier
\vspace{-1em}
